---
title: "Coursera Regression Assignment"
author: "David Smith"
date: "Thursday, July 17, 2014"
output: word_document
---

Executive Summary
==================
Miles per gallon (MPG) as a function of transmission type depends upon vehicle weight.  In general, vehicles with a manual transmission weigh less and get more MPG than vehicles with an automatic transmission.  However, due to the interaction between vehicle weight and transmission, heavy vehicles (>3000 Lbs) with a manual transmission are likely to get worse gas mileage than heavy vehicles (>3000 Lbs) with an automatic transmission.  

Specifically, the expected difference in gas mileage between manual and automatic transmission cars at light, medium, and high vehicle weights are shown below:  

* Light (2000 Lbs):     Manual ~ 4.3 MPG Better Than Automatic
* Medium (3000 Lbs):    Manual MPG and Automatic MPG Approximately Equal
* Heavy (4000 Lbs):     Automatic ~ 6.3 MPG Better Than Manual

In addition to the predicted difference in gas mileage, the *relationship* between gas mileage and vehicle weight for each type of transmission are as follows:

* Manual Transmissions:      Lose ~ 9.1 MPG for every additional 1000 LBS
* Automatic Transmissions:   Lose ~ 3.8 MPG for every additional 1000 LBS

Method
========
* Explore with pairs plot of all bivariate relationships between variables (Appendix)
* Explore full model with all explanatory variables to ID potentially important variables
* Sequentially fit nested models (with anova) and compare for importance / significance
* Run diagnostics on "best" model - diagnostic plots, dffits, dfbetas, hatvalues, PRESS
* Calculate confidence intervals for coefficients and predictions (at specified weights)


```{r GetAndCleanData, echo=FALSE, results="hide"}
# Get and Clean Data
data(mtcars)

names(mtcars)<-c("MPG","Cylinders","Displacement","Horsepower","Rear.Axle.Ratio","Weight","Quarter.Mile.Time","VS","Transmission","Gears","Carburetors")

mtcars$Transmission<-factor(mtcars$Transmission,levels=c(0,1),labels=c("Automatic","Manual"))

mtcars$VS<-factor(mtcars$VS,levels=c(0,1))
head(mtcars)
```


Results
=========
```{r BestModel, echo=FALSE}
# Determined By Evaluating Nested Models (See Appendix)
myFit3<- lm(MPG ~ Weight + Transmission + Weight*Transmission, data = mtcars)
# Get R-Sq Value
myRSq2<-round(100*summary(myFit3)$r.squared,1)
out<-summary(myFit3)

```
The final, parsimonious, best-fitting model includes terms for the weight, type of transmission, and the interaction between weight and transmission.  It accounts for `r myRSq2`% of the variation in MPG.  The figure below visually summarizes the relationship between MPG and weight for each type of transmission.
```{r Plot1, echo=FALSE, fig.height=3.7}
# Make Subsets
mtcarsAuto<-mtcars[mtcars$Transmission=="Automatic",]
mtcarsManual<-mtcars[mtcars$Transmission=="Manual",]
 
# Fit Subsets
autoFit<-lm(MPG ~ Weight, data=mtcarsAuto)
manualFit<-lm(MPG ~ Weight, data=mtcarsManual)

# Plot
par(mfrow=c(1,1))
plot(mtcars$Weight,mtcars$MPG,col=mtcars$Transmission,xlim=c(0,6),ylim=c(0,50),xlab="Weight (1000/LBS)",ylab="Miles Per Gallon (MPG)")
abline(autoFit,col="black")
abline(manualFit,col="red")

text(x=4.5,y=40,labels=c("Manual = Red"),col="red",cex=0.75,pos=4)
text(x=4.5,y=37,labels=c("Automatic = Black"),col="black",cex=0.75,pos=4)

```
Predictions and their corresponding uncertainty are given below at three different weights for each type of transmission.  
```{r Predictions, echo=FALSE}
new<-data.frame(Transmission=c(rep("Manual",3),rep("Automatic",3)),
                Weight=as.numeric(c(rep(c("2","3","4"),2))))

myPredict<-cbind(new,as.data.frame(round(predict(myFit3,new,interval="confidence"),1)))
names(myPredict)<-c("Transmission","Weight","Estimated MPG","Lower 95% Level","Upper 95% Level")
myPredict$Weight<-myPredict$Weight*1000
myPredict

```

Note that there is little to no overalap in the 95% confidence intervals between the manual and automatic transmissions at 2000 LBS and 4000 LBS, suggesting that there is a significant difference in MPG at these weights.  At 3000 LBS, the 95% confidence intervals largely overlap, indicating that there is no significant difference in MPG at this weight.  

Confidence intervals for the regression coefficients are also provided (below).  
```{r BetasCI, echo=FALSE}
myBetas<-cbind(data.frame(round(myFit3$coefficients,1)),round(confint(myFit3),1))
names(myBetas)<-c("Estimate","Lower 95% Level","Upper 95% Level")
myBetas
```

The intercept for automatic transmissions is ~ 31.4 MPG (at the nonsensical 0 LBS), while the intercept for manual transmissions is ~ 14.9 MPG higher, or 46.3 MPG.  The slope for automatic transmissions is approximately -3.8 MPG per 1000 LBS, while it is about -9.1 MPG per additional 1000 LBS. Diagnostics for this model are shown in the appendix.


Appendix
=========
 

```{r Explore, echo=FALSE, fig.height=6,  fig.width=6}
# From help(cor)
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
pairs(mtcars, lower.panel=panel.cor,upper.panel=panel.smooth)

```



```{r Diagnostics2, echo=FALSE, fig.width=6,fig.height=6}

par(mfrow=c(2,2))
plot(myFit3)


```


